---
title: "Process Data for ML"
author: "Matthew Berginski"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(here)
library(tidymodels)

knitr::opts_knit$set(root.dir = here())
```

The following section reads in the klaeger data and filters to the collection of compounds that are also in the synergy screen. The data is also rearranged to match up with the expectation that the relative intensity values for every gene will be in a seperate column.

```{r}
drug_matches = read_csv(here('src/find_screen_klaeger_matches/klaeger_synergy_drug_matches.csv'))

klaeger = read_rds(here('results/klaeger_full_tidy.rds')) %>%
	mutate(drug = trimws(drug))

klaeger_compound_matches = klaeger %>%
	#filter to just the drugs that are in the match list and also toss out the zero
	#concentration values
	filter(drug %in% drug_matches$klaeger_drugs, concentration_M != 0)

#Make sure all the expected matches are found in the filtered Klaeger set
assertthat::assert_that(all(klaeger_compound_matches$drug %in% drug_matches$klaeger_drugs))

klaeger_wide = klaeger_compound_matches %>%
	pivot_wider(names_from = "gene_name", values_from = "relative_intensity")
```

The synergy data is processed in much the same way, toss out all the compounds not in the match list and prep the data for combining with Klaeger.

```{r}
synergy = read_rds(here('results/synergy_combined.rds')) %>%
	mutate(anchor = trimws(anchor), compound = trimws(compound))

synergy_compound_matches = bind_rows(
	synergy %>%
		filter(compound %in% drug_matches$synergy_drugs, dose_anchor_m == 0, dose_compound_m != 0),
	synergy %>%
		filter(anchor %in% drug_matches$synergy_drugs, dose_compound_m == 0, dose_anchor_m != 0)
) %>% mutate(drug = case_when(
	dose_anchor_m == 0 ~ compound,
	dose_compound_m == 0 ~ anchor
)) %>% mutate(dose = case_when(
	dose_anchor_m == 0 ~ dose_compound_m,
	dose_compound_m == 0 ~ dose_anchor_m
))

#Make sure all the expected matches are found in the filtered Klaeger set
assertthat::assert_that(all(synergy_compound_matches$drug %in% drug_matches$synergy_drugs))

synergy_compound_to_join = synergy_compound_matches %>%
	select(drug, dose, cell_line, viability) %>%
	left_join(drug_matches, by=c('drug' = 'synergy_drugs')) %>%
	select(-drug)
```
Now we join up the synergy and Klaeger, followed by a filter to toss out any gene that doesn't vary at all across the matched compounds. Then save out the data frame with the raw viability values and produce a second data set with the viability converted into a binary above/below 90 threshold.

```{r}
klaeger_synergy = klaeger_wide %>%
	left_join(synergy_compound_to_join, by=c('drug'='klaeger_drugs','concentration_M'='dose')) %>%
	select(drug,concentration_M,viability,cell_line,everything()) %>%
	filter(!is.na(viability)) %>% 
	arrange(cell_line,drug,concentration_M)

klaeger_synergy_tidy = klaeger_synergy %>%
	pivot_longer(-c('drug','concentration_M','viability','cell_line'), names_to = "gene_name", values_to = 'relative_intensity')

no_gene_variation = klaeger_synergy_tidy %>%
	group_by(gene_name) %>%
	summarise(gene_sd = sd(relative_intensity)) %>%
	filter(gene_sd == 0)

klaeger_synergy_gene_variation = klaeger_synergy %>%
	#toss out the genes with no variation
	select(-one_of(no_gene_variation$gene_name))

write_rds(klaeger_synergy_gene_variation,here('results/klaeger_screen_for_regression.rds'), compress='gz')

klaeger_synergy_below_90 = klaeger_synergy_gene_variation %>%
	mutate(viability_90 = viability <= 90) %>%
	select(-viability) %>%
	select(drug,concentration_M,viability_90,cell_line,everything()) %>% 
	write_rds(here('results/klaeger_synergy_for_classification_90.rds'), compress='gz')

klaeger_for_prediction = klaeger %>% 
	filter(!drug %in% drug_matches$klaeger_drugs, 
				 concentration_M != 0,
				 ! gene_name %in% no_gene_variation$gene_name)

#After filtering out all the genes without variance in the klaeger set, there
#are probably a few drugs that don't hit any of the genes included in the model.
#So I'll find those and filter them in the next step.
klaeger_prediction_no_variation = klaeger_for_prediction %>% 
	group_by(drug) %>% 
	summarise(relative_sd = sd(relative_intensity)) %>% 
	filter(relative_sd == 0)

klaeger_wide_for_prediction = klaeger_for_prediction %>%
	filter(! drug %in% klaeger_prediction_no_variation$drug) %>%
	pivot_wider(names_from = "gene_name", values_from = "relative_intensity") %>%
	write_rds(here('results/klaeger_wide_for_prediction.rds'), compress='gz')
```

Finally, in prep for running cross validation across leave-one-compound-out:

```{r}
cell_line_compound_splits = list()

for (this_cell_line in unique(klaeger_synergy_below_90$cell_line)) {
	
	klaeger_data_cell_line = klaeger_synergy_below_90 %>%
		filter(cell_line == this_cell_line)
	
	klaeger_data_cell_line = klaeger_data_cell_line %>%
		select(-cell_line,-concentration_M) %>%
		mutate(viability_90 = as.factor(viability_90))
	
	splits = list()
	index = 1
	id = c()
	for (exclude_compound in unique(klaeger_data_cell_line$drug)) {
		assessment_ids = which(klaeger_data_cell_line$drug == exclude_compound)
		analysis_ids = which(klaeger_data_cell_line$drug != exclude_compound)
		
		splits[[index]] = make_splits(list("analysis" = analysis_ids,"assessment" = assessment_ids),
																	klaeger_data_cell_line %>% select(-drug))
		index = index + 1
		
		id = c(id,exclude_compound)
	}
	
	cell_line_compound_splits[[this_cell_line]] = new_rset(
		splits = splits,
		ids = id,
		attrib = paste0("Per compound cv splits for ", this_cell_line),
		subclass = c("vfold_cv", "rset")
	)
}

write_rds(cell_line_compound_splits, here('results/klaeger_synergy_classification_90_CV_split.rds'), compress = 'gz')
```

```{r}
cell_line_compound_splits = list()

for (this_cell_line in unique(klaeger_synergy_gene_variation$cell_line)) {
	
	klaeger_data_cell_line = klaeger_synergy_gene_variation %>%
		filter(cell_line == this_cell_line) %>%
		select(-cell_line,-concentration_M)
	
	splits = list()
	index = 1
	id = c()
	for (exclude_compound in unique(klaeger_data_cell_line$drug)) {
		assessment_ids = which(klaeger_data_cell_line$drug == exclude_compound)
		analysis_ids = which(klaeger_data_cell_line$drug != exclude_compound)
		
		splits[[index]] = make_splits(list("analysis" = analysis_ids,"assessment" = assessment_ids),
																	klaeger_data_cell_line %>% select(-drug))
		index = index + 1
		
		id = c(id,exclude_compound)
	}
	
	cell_line_compound_splits[[this_cell_line]] = new_rset(
		splits = splits,
		ids = id,
		attrib = paste0("Per compound cv splits for ", this_cell_line),
		subclass = c("vfold_cv", "rset")
	)
	print(names(cell_line_compound_splits))
}

write_rds(cell_line_compound_splits, here('results/klaeger_synergy_regression_CV_split.rds'), compress = 'gz')
```