---
title: "Build Klaeger PDAC Synergy Model - Random Forest"
author: "Matthew Berginski"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
library(tidymodels)
library(tidyverse)
library(tictoc)
library(here)
library(patchwork)
library(gt)
library(ggpubr)
library(MESS)
library(gridExtra)
library(vip)

# tidymodels
library(rsample)
library(recipes)
library(parsnip)
library(tune)
library(dials)
library(workflows)
library(yardstick)

# speed up computation with parallel processing (optional)
library(doParallel)
all_cores <- parallel::detectCores(logical = FALSE)
doParallel::registerDoParallel(cores = all_cores - 2)

knitr::opts_knit$set(root.dir = here::here())
```

# Read In and Combine Klaeger/Synergy Data

This chunk of code reads in the pre-processed Klaeger kinase activation data and the combined synergy screen data. These two data sets are matched up and and then used in the rest of the model building steps.

```{r, include=F}
klaeger_data = read_rds(here('data/klaeger_full_tidy.rds'))

klaeger_data_wide = klaeger_data %>%
	pivot_wider(names_from = gene_name, values_from = relative_intensity)

```

```{r combine synergy and klaeger, include=F}
synergy_data_only_single = read_rds(here('data/synergy_combined.rds')) %>%
	mutate(anchor_lower = tolower(anchor),
				 compound_lower = tolower(compound)) %>%
	filter(dose_anchor_m == 0 | dose_compound_m == 0) %>%
	#The NAF cell line wasn't correctly labeled, so remove from data
	filter(cell_line != "NAF")

compound_matches = read_csv(here('data/DrugMatches.csv')) %>%
	#removing one of the alisertib compound matches that aren't in my copy of the synergy
	filter(pdac != "mln8237")

klaeger_data_matches = klaeger_data_wide %>%
	#All of the Klaeger data with zero compound concentration are 1 across the
	#board, also I have no particular way to match up the the no-treatments with
	#all the no-treatments from the synergy screen.
	filter(concentration_M != 0) %>%
	mutate(drug_lower = tolower(drug)) %>%
	filter(drug_lower %in% compound_matches$klaeger) %>%
	left_join(compound_matches, by=c('drug_lower'='klaeger')) %>%
	select(drug,drug_lower,pdac,everything())

klaeger_data_matches_compound = klaeger_data_matches %>% 
	left_join(synergy_data_only_single %>% 
							filter(dose_anchor_m == 0) %>%
							select(compound_lower,dose_compound_m,viability,cell_line), 
						by=c('pdac' = 'compound_lower', 'concentration_M' = 'dose_compound_m')) %>%
	select(drug,drug_lower,pdac,cell_line,viability,everything()) %>%
	filter(!is.na(viability))

klaeger_data_matches_anchor = klaeger_data_matches %>% 
	left_join(synergy_data_only_single %>% 
							filter(dose_compound_m == 0) %>%
							select(anchor_lower,dose_anchor_m,viability,cell_line), 
						by=c('pdac' = 'anchor_lower', 'concentration_M' = 'dose_anchor_m')) %>%
	select(drug,drug_lower,pdac,cell_line,viability,everything()) %>%
	filter(!is.na(viability))

klaeger_data_matches_full = bind_rows(
	klaeger_data_matches_compound,
	klaeger_data_matches_anchor) %>% 
	mutate(viability_binary = as.factor(viability < 90)) %>%
	select(drug,drug_lower,pdac,cell_line,viability,viability_binary,everything())

klaeger_data_matches_tidy = klaeger_data %>% 
	filter(drug %in% unique(klaeger_data_matches_full$drug), 
				 concentration_M %in% unique(klaeger_data_matches_full$concentration_M))

no_gene_variation = klaeger_data_matches_tidy %>%
	group_by(gene_name) %>%
	summarise(gene_sd = sd(relative_intensity)) %>%
	filter(gene_sd == 0)

klaeger_data_matches_full = klaeger_data_matches_full %>%
	select(-one_of(no_gene_variation$gene_name))
```

```{r, include=F}
assess_threshold = function(threshold,data) {
	data %>% 
		mutate(actual_viability = as.logical(actual_viability_binary)) %>%
		mutate(above_thresh = predicted_viability >= threshold) %>% 
		summarise(cutoff = threshold,
							TN = sum(! actual_viability & ! above_thresh),
							FP = sum(! actual_viability & above_thresh),
							FN = sum(actual_viability & ! above_thresh),
							TP = sum(actual_viability & above_thresh))
}

gather_ROC_vals = function(input_data, prediction_type, sample_points = 50) {
	ROC_vals = quantile(input_data$predicted_viability,seq(0,1,length.out = sample_points)) %>%
		map(~ assess_threshold(., data=input_data)) %>%
		reduce(rbind) %>%
		mutate(TPR = TP/(TP+FN), 
					 FPR = FP/(FP+TN),
					 precision = TP/(TP+FP),
					 recall = TP/(TP+FN)) %>%
		mutate(prediction_type = prediction_type) %>%
		identity()
	return(ROC_vals)
}
```

```{r distribution of viability}
ggplot(klaeger_data_matches_full, aes(x = viability)) + 
	geom_histogram() +
	labs(x="Cell Viability", y="Number of Drug/Cell Line Combos") +
	geom_vline(aes(xintercept=90), color='red') +
	BerginskiRMisc::theme_berginski()
```

# Modeling

The modeling results are divided into two parts, one model that attempts to predict the likelihood that the cell viability is less than 90 and another that tries to predict cell viability below 40. I picked below 90 and below 40 because the data is split in half at below 90 and 10% of the data is below 40. Each cell line is modeled independently, but I've combined the resulting ROC/PRC curves below. I've built models for cross validation ROC/PRC curves for both sets of models with several different cross validation stratifications:

* Leave One Compound Out - all of a single compounds data (all concentrations) are left out for prediction
* Leave One Compound Concentration Out - all of a given compound's concentration values are left out for prediction (each given compound concentration was tested about 8 times in the data set)
* Leave One Measurment Out - only a single measurement was excluded for prediction (this isn't a very good test at all, given that there are ~8 repeats of the same experiment in the data, but I've included anyway for the sake of comparison)

## Boost Tree - Below 90 Predictions - Leave One Compound Out

```{r, cache=T, include=F}
tic()

prediction_results = c()
for (this_cell_line in unique(klaeger_data_matches_full$cell_line)) {
	
	klaeger_data_cell_line = klaeger_data_matches_full %>%
		filter(cell_line == this_cell_line)
	
	for (exclude_compound in unique(klaeger_data_cell_line$drug)) {
		
		klaeger_train = klaeger_data_cell_line %>%
			filter(drug != exclude_compound) %>%
			select(-drug,-drug_lower,-pdac,-concentration_M,-cell_line,-viability) 
		
		klaeger_test = klaeger_data_cell_line %>%
			filter(drug == exclude_compound) %>%
			select(-drug,-drug_lower,-pdac,-concentration_M,-cell_line,-viability_binary,-viability)
		
		these_prediction_results = klaeger_data_cell_line %>%
			filter(drug == exclude_compound) %>%
			select(drug,drug_lower,pdac,concentration_M,cell_line,viability_binary,viability) %>%
			rename(actual_viability_binary = viability_binary, actual_viability = viability)
		
		xg_mod <-
			boost_tree(trees = 1850,
								 mtry = 1,
								 min_n = 13,
								 tree_depth = 2,
								 learn_rate = 0.007338206,
								 loss_reduction = 1.569039e-06,
								 sample_size = 0.5074992) %>%
			set_engine("xgboost") %>%
			set_mode("classification")
		
		xg_fit_rs <-
			xg_mod %>%
			fit(viability_binary ~ ., data = klaeger_train)
		
		these_prediction_results$predicted_viability = predict(xg_fit_rs, klaeger_test, type = "prob")$.pred_TRUE
		these_prediction_results$test_set_size = dim(klaeger_test)[1]
		
		prediction_results = bind_rows(
			prediction_results,
			these_prediction_results
		)
	}
}

toc()
```

```{r, include=F}
ROC_vals = bind_rows(
	gather_ROC_vals(prediction_results %>% filter(cell_line == "CAF"), "CAF"),
	gather_ROC_vals(prediction_results %>% filter(cell_line == "P1004"), "P1004"),
	gather_ROC_vals(prediction_results %>% filter(cell_line == "P1304"), "P1304"),
) %>% rename(cell_line = prediction_type) %>%
	arrange(cell_line,desc(cutoff))

ROC_measurements = ROC_vals %>% 
	group_by(cell_line) %>% 
	summarise("ROC_AUC" = signif(auc(x = 1 - FPR, y = TPR),3),
						"PRC_AUC" = signif(auc(x = recall, y = precision),3)) %>%
	ungroup() %>%
	mutate("AUC_rank" = min_rank(ROC_AUC),
				 "PRC_rank" = min_rank(PRC_AUC),
				 "sum_rank" = AUC_rank + PRC_rank) %>%
	arrange(desc(sum_rank)) %>%
	select(-contains('rank'))
```

```{r below 90 1000 tree,fig.height=5,fig.width=12, echo=F}
null_line = data.frame(x1 = 0, x2 = 1, y1 = 0, y2 = 1)

AUC_plot = ggplot(ROC_vals) + 
	geom_segment(data=null_line, mapping=aes(x=x1,y=y1,xend=x2,yend=y2),alpha=0.25) +
	# geom_point(aes(x=FPR,y=TPR, color=cell_line),lwd=1) + 
	geom_line(aes(x=FPR,y=TPR, color=cell_line),lwd=1) +
	labs(x="FPR",y="TPR") +
	theme(legend.position = "none") +
	# ggtitle("Below 90 LOO Drug") +
	BerginskiRMisc::theme_berginski()

null_line = data.frame(x1 = 0, x2 = 1, 
											 y1 = mean(as.logical(klaeger_data_matches_full$viability_binary)), 
											 y2 = mean(as.logical(klaeger_data_matches_full$viability_binary)))

PRC_plot = ggplot(ROC_vals) + 
	geom_segment(data=null_line, mapping=aes(x=x1,y=y1,xend=x2,yend=y2),alpha=0.25) +
	# geom_point(aes(x=FPR,y=TPR, color=cell_line),lwd=1) + 
	geom_line(aes(x=recall,y=precision, color=cell_line),lwd=1) +
	labs(x="Recall",y="Precision", color="Cell Line") +
	# xlim(c(0.075,0.25)) +
	BerginskiRMisc::theme_berginski()

ROC_table = tableGrob(ROC_measurements, row = rep("",dim(ROC_measurements)[1]))

full_plots = ggarrange(AUC_plot, PRC_plot, ROC_table,nrow=1)
full_plots
```

## Boost Tree - Below 90 Predictions - Leave One Compound Out

```{r, cache=T, include=F}
tic()

doParallel::registerDoParallel()

prediction_results = c()
for (this_cell_line in unique(klaeger_data_matches_full$cell_line)) {
	
	klaeger_data_cell_line = klaeger_data_matches_full %>%
		filter(cell_line == this_cell_line)
	
	for (exclude_compound in unique(klaeger_data_cell_line$drug)) {
		
		klaeger_train = klaeger_data_cell_line %>%
			filter(drug != exclude_compound) %>%
			select(-drug,-drug_lower,-pdac,-concentration_M,-cell_line,-viability) 
		
		klaeger_test = klaeger_data_cell_line %>%
			filter(drug == exclude_compound) %>%
			select(-drug,-drug_lower,-pdac,-concentration_M,-cell_line,-viability_binary,-viability)
		
		these_prediction_results = klaeger_data_cell_line %>%
			filter(drug == exclude_compound) %>%
			select(drug,drug_lower,pdac,concentration_M,cell_line,viability_binary,viability) %>%
			rename(actual_viability_binary = viability_binary, actual_viability = viability)
		
		xg_mod <-
			boost_tree(trees = 1000, tree_depth = 10) %>%
			set_engine("xgboost") %>%
			set_mode("classification")
		
		xg_fit_rs <-
			xg_mod %>%
			fit(viability_binary ~ ., data = klaeger_train)
		
		these_prediction_results$predicted_viability = predict(xg_fit_rs, klaeger_test, type = "prob")$.pred_TRUE
		these_prediction_results$test_set_size = dim(klaeger_test)[1]
		
		prediction_results = bind_rows(
			prediction_results,
			these_prediction_results
		)
	}
}

toc()
```

```{r, include=F}
ROC_vals = bind_rows(
	gather_ROC_vals(prediction_results %>% filter(cell_line == "CAF"), "CAF"),
	gather_ROC_vals(prediction_results %>% filter(cell_line == "P1004"), "P1004"),
	gather_ROC_vals(prediction_results %>% filter(cell_line == "P1304"), "P1304"),
) %>% rename(cell_line = prediction_type) %>%
	arrange(cell_line,desc(cutoff))

ROC_measurements = ROC_vals %>% 
	group_by(cell_line) %>% 
	summarise("ROC_AUC" = signif(auc(x = 1 - FPR, y = TPR),3),
						"PRC_AUC" = signif(auc(x = recall, y = precision),3)) %>%
	ungroup() %>%
	mutate("AUC_rank" = min_rank(ROC_AUC),
				 "PRC_rank" = min_rank(PRC_AUC),
				 "sum_rank" = AUC_rank + PRC_rank) %>%
	arrange(desc(sum_rank)) %>%
	select(-contains('rank'))
```

```{r below 90 1000 tree,fig.height=5,fig.width=12, echo=F}
null_line = data.frame(x1 = 0, x2 = 1, y1 = 0, y2 = 1)

AUC_plot = ggplot(ROC_vals) + 
	geom_segment(data=null_line, mapping=aes(x=x1,y=y1,xend=x2,yend=y2),alpha=0.25) +
	# geom_point(aes(x=FPR,y=TPR, color=cell_line),lwd=1) + 
	geom_line(aes(x=FPR,y=TPR, color=cell_line),lwd=1) +
	labs(x="FPR",y="TPR") +
	theme(legend.position = "none") +
	# ggtitle("Below 90 LOO Drug") +
	BerginskiRMisc::theme_berginski()

null_line = data.frame(x1 = 0, x2 = 1, 
											 y1 = mean(as.logical(klaeger_data_matches_full$viability_binary)), 
											 y2 = mean(as.logical(klaeger_data_matches_full$viability_binary)))

PRC_plot = ggplot(ROC_vals) + 
	geom_segment(data=null_line, mapping=aes(x=x1,y=y1,xend=x2,yend=y2),alpha=0.25) +
	# geom_point(aes(x=FPR,y=TPR, color=cell_line),lwd=1) + 
	geom_line(aes(x=recall,y=precision, color=cell_line),lwd=1) +
	labs(x="Recall",y="Precision", color="Cell Line") +
	# xlim(c(0.075,0.25)) +
	BerginskiRMisc::theme_berginski()

ROC_table = tableGrob(ROC_measurements, row = rep("",dim(ROC_measurements)[1]))

full_plots = ggarrange(AUC_plot, PRC_plot, ROC_table,nrow=1)
full_plots
```
