---
title: "Prep Validation Screen Data"
author: "Matthew Berginski"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(here)
library(readxl)
library(patchwork)

knitr::opts_knit$set(root.dir = here())
```

# Read in plate data

All the data from the plate reader is collected into a set of excel spreadsheets. The luminescence values show up at the bottom of the spreadsheet. I've manually matched up the names of the excel files to the appropriate cell line names and compounds in each plate. There is a text file with the same information in the same data folder I've read these files from.

The plate is organized with the concentrations decreasing over each column of the plate, following the concentration used in Klaeger. The last two columns in each plate are DMSO controls.

```{r}
read_xlsx_plate_data <- function(file_name,these_compounds,this_cell_line,skip_lines = 16) {
	
	temp_plate = read_excel(file_name, skip = skip_lines) %>%
		select(-A, -`...2`) %>%
		slice(1:6) %>%
		mutate(row = c(seq(6))) %>% 
		mutate(compound = c(rep(these_compounds[1],3),rep(these_compounds[2],3))) %>%
		pivot_longer(-c(compound,row), names_to = "concentration_M",values_to = "lum") %>%
		mutate(cell_line = this_cell_line, plate_name = basename(file_name)) %>%
		identity()
	
	DMSO_vals = temp_plate %>%
		filter(concentration_M == "...11" | concentration_M == "...12") %>%
		mutate(concentration_M = 0) %>%
		mutate(compound = "DMSO")
	
	temp_plate = temp_plate %>%
		filter(concentration_M != "...11", concentration_M != "...12") %>%
		mutate(concentration_M = case_when(
			concentration_M == "...3" ~ 3e-5,
			concentration_M == "...4" ~ 3e-6,
			concentration_M == "...5" ~ 1e-6,
			concentration_M == "...6" ~ 3e-7,
			concentration_M == "...7" ~ 1e-7,
			concentration_M == "...8" ~ 3e-8,
			concentration_M == "...9" ~ 1e-8,
			concentration_M == "...10" ~ 3e-9,
			T ~ NA_real_
		)) %>%
		bind_rows(DMSO_vals)
	
	return(temp_plate)
}

read_csv_plate_data <- function(file_name,these_compounds,this_cell_line) {
	
	temp_plate = read_csv(file_name, col_names = FALSE, show_col_types = F) %>%
		select(-X1, -X12) %>%
		slice(2:7) %>%
		mutate(row = 1:6) %>% 
		mutate(compound = c(rep(these_compounds[1],3),rep(these_compounds[2],3))) %>%
		pivot_longer(-c(compound,row), names_to = "concentration_M",values_to = "lum") %>%
		mutate(cell_line = this_cell_line, plate_name = basename(file_name)) %>%
		mutate(lum = as.numeric(lum)) %>% 
		identity()
	
	DMSO_vals = temp_plate %>%
		filter(concentration_M == "X10" | concentration_M == "X11") %>%
		mutate(concentration_M = 0) %>%
		mutate(compound = "DMSO")
	
	temp_plate = temp_plate %>%
		filter(concentration_M != "X10", concentration_M != "X11") %>%
		mutate(concentration_M = case_when(
			concentration_M == "X2" ~ 3e-5,
			concentration_M == "X3" ~ 3e-6,
			concentration_M == "X4" ~ 1e-6,
			concentration_M == "X5" ~ 3e-7,
			concentration_M == "X6" ~ 1e-7,
			concentration_M == "X7" ~ 3e-8,
			concentration_M == "X8" ~ 1e-8,
			concentration_M == "X9" ~ 3e-9,
			T ~ NA_real_
		)) %>%
		bind_rows(DMSO_vals)
	
	return(temp_plate)
}

plate_vals_part3 = bind_rows(
	read_csv_plate_data(here('data/PDAC_validation_screen_part3/119_1.csv'),c("Masitinib","Ripasudil"),"CAF"),
	read_csv_plate_data(here('data/PDAC_validation_screen_part3/119_2.csv'),c("AT-13148","RGB-286638"),"CAF"),
	read_csv_plate_data(here('data/PDAC_validation_screen_part3/119_3.csv'),c("PHA-793887","AT-9283"),"CAF"),
	read_csv_plate_data(here('data/PDAC_validation_screen_part3/119_4.csv'),c("Lestaurtinib","KW-2449"),"CAF"),
	read_csv_plate_data(here('data/PDAC_validation_screen_part3/119_5.csv'),c("K-252a","PF-03814735"),"CAF"),
	read_csv_plate_data(here('data/PDAC_validation_screen_part3/119_6.csv'),c("XL-228","JANEX-1"),"CAF"),
	
	read_csv_plate_data(here('data/PDAC_validation_screen_part3/411_1.csv'),c("Masitinib","Ripasudil"),"P1304"),
	read_csv_plate_data(here('data/PDAC_validation_screen_part3/411_2.csv'),c("AT-13148","RGB-286638"),"P1304"),
	read_csv_plate_data(here('data/PDAC_validation_screen_part3/411_3.csv'),c("PHA-793887","AT-9283"),"P1304"),
	read_csv_plate_data(here('data/PDAC_validation_screen_part3/411_4.csv'),c("Lestaurtinib","KW-2449"),"P1304"),
	read_csv_plate_data(here('data/PDAC_validation_screen_part3/411_5.csv'),c("K-252a","PF-03814735"),"P1304"),
	read_csv_plate_data(here('data/PDAC_validation_screen_part3/411_6.csv'),c("XL-228","JANEX-1"),"P1304"),
	
	read_csv_plate_data(here('data/PDAC_validation_screen_part3/422_1.csv'),c("Masitinib","Ripasudil"),"P1004"),
	read_csv_plate_data(here('data/PDAC_validation_screen_part3/422_2.csv'),c("AT-13148","RGB-286638"),"P1004"),
	read_csv_plate_data(here('data/PDAC_validation_screen_part3/422_3.csv'),c("PHA-793887","AT-9283"),"P1004"),
	read_csv_plate_data(here('data/PDAC_validation_screen_part3/422_4.csv'),c("Lestaurtinib","KW-2449"),"P1004"),
	read_csv_plate_data(here('data/PDAC_validation_screen_part3/422_5.csv'),c("K-252a","PF-03814735"),"P1004"),
	read_csv_plate_data(here('data/PDAC_validation_screen_part3/422_6.csv'),c("XL-228","JANEX-1"),"P1004")
)

plate_vals_part2 = bind_rows(
	read_xlsx_plate_data(here('data/PDAC_validation_screen_part2/119_1.xlsx'),c("Masitinib","Ripasudil"),"CAF"),
	read_xlsx_plate_data(here('data/PDAC_validation_screen_part2/119_2.xlsx'),c("AT-13148","RGB-286638"),"CAF"),
	read_xlsx_plate_data(here('data/PDAC_validation_screen_part2/119_3.xlsx'),c("PHA-793887","AT-9283"),"CAF"),
	read_xlsx_plate_data(here('data/PDAC_validation_screen_part2/119_4.xlsx'),c("Lestaurtinib","KW-2449"),"CAF"),
	read_xlsx_plate_data(here('data/PDAC_validation_screen_part2/119_5.xlsx'),c("K-252a","PF-03814735"),"CAF"),
	read_xlsx_plate_data(here('data/PDAC_validation_screen_part2/119_6.xlsx'),c("XL-228","JANEX-1"),"CAF"),
	
	read_xlsx_plate_data(here('data/PDAC_validation_screen_part2/411_1.xlsx'),c("Masitinib","Ripasudil"),"P1304"),
	read_xlsx_plate_data(here('data/PDAC_validation_screen_part2/411_2.xlsx'),c("AT-13148","RGB-286638"),"P1304"),
	read_xlsx_plate_data(here('data/PDAC_validation_screen_part2/411_3.xlsx'),c("PHA-793887","AT-9283"),"P1304"),
	read_xlsx_plate_data(here('data/PDAC_validation_screen_part2/411_4.xlsx'),c("Lestaurtinib","KW-2449"),"P1304"),
	read_xlsx_plate_data(here('data/PDAC_validation_screen_part2/411_5.xlsx'),c("K-252a","PF-03814735"),"P1304"),
	read_xlsx_plate_data(here('data/PDAC_validation_screen_part2/411_6.xlsx'),c("XL-228","JANEX-1"),"P1304"),
	
	read_xlsx_plate_data(here('data/PDAC_validation_screen_part2/422_1.xlsx'),c("Masitinib","Ripasudil"),"P1004"),
	read_xlsx_plate_data(here('data/PDAC_validation_screen_part2/422_2.xlsx'),c("AT-13148","RGB-286638"),"P1004"),
	read_xlsx_plate_data(here('data/PDAC_validation_screen_part2/422_3.xlsx'),c("PHA-793887","AT-9283"),"P1004"),
	read_xlsx_plate_data(here('data/PDAC_validation_screen_part2/422_4.xlsx'),c("Lestaurtinib","KW-2449"),"P1004"),
	read_xlsx_plate_data(here('data/PDAC_validation_screen_part2/422_5.xlsx'),c("K-252a","PF-03814735"),"P1004"),
	read_xlsx_plate_data(here('data/PDAC_validation_screen_part2/422_6.xlsx'),c("XL-228","JANEX-1"),"P1004")
)

plate_vals_part1 = bind_rows(
	read_xlsx_plate_data(here('data/PDAC_validation_screen_part1/411_1.xlsx'),c("Masitinib","Ripasudil"),"P1304", skip_lines = 14),
	read_xlsx_plate_data(here('data/PDAC_validation_screen_part1/411_2.xlsx'),c("AT-13148","RGB-286638"),"P1304", skip_lines = 14),
	read_xlsx_plate_data(here('data/PDAC_validation_screen_part1/411_3.xlsx'),c("PHA-793887","AT-9283"),"P1304", skip_lines = 14),
	read_xlsx_plate_data(here('data/PDAC_validation_screen_part1/411_4.xlsx'),c("Lestaurtinib","KW-2449"),"P1304", skip_lines = 14),
	read_xlsx_plate_data(here('data/PDAC_validation_screen_part1/411_5a.xlsx'),c("PF-03814735","XL-228"),"P1304", skip_lines = 14),
	
	read_xlsx_plate_data(here('data/PDAC_validation_screen_part1/422_1.xlsx'),c("Masitinib","Ripasudil"),"P1004", skip_lines = 14),
	read_xlsx_plate_data(here('data/PDAC_validation_screen_part1/422_2.xlsx'),c("AT-13148","RGB-286638"),"P1004", skip_lines = 14),
	read_xlsx_plate_data(here('data/PDAC_validation_screen_part1/422_3.xlsx'),c("PHA-793887","AT-9283"),"P1004", skip_lines = 14),
	read_xlsx_plate_data(here('data/PDAC_validation_screen_part1/422_4.xlsx'),c("Lestaurtinib","KW-2449"),"P1004", skip_lines = 14),
	read_xlsx_plate_data(here('data/PDAC_validation_screen_part1/422_5a.xlsx'),c("PF-03814735","XL-228"),"P1004", skip_lines = 14)
)

if (any(is.na(plate_vals_part2$concentration_M)) | any(is.na(plate_vals_part3$concentration_M))) {
	stop("Found missing concentration value, this shouldn't happen.")
}

```

# Per Plate DMSO Boxplots

```{r}
temp = plate_vals_part3 %>%
	filter(compound == "DMSO") %>%
	bind_rows(
		plate_vals_part2 %>%
			filter(compound == "DMSO") %>%
			mutate(plate_name = paste0("p1_",plate_name))
		
	)

temp = temp %>%
	bind_rows(
		plate_vals_part1 %>%
			filter(compound == "DMSO") %>%
			mutate(plate_name = paste0("p1_",plate_name))
	)


ggplot(temp, aes(x=plate_name,y=lum)) + 
	geom_boxplot() +
	# ylim(c(min(plate_vals$lum), NA)) +
	theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```


```{r}
treatment_lum_summaries_part2 = plate_vals_part2 %>%
	filter(compound != "DMSO") %>% 
	group_by(row, compound, concentration_M, cell_line) %>%
	summarise(sd = sd(lum),
						num_samples = n(),
						mean_lum = mean(lum),
						CV = sd/mean_lum)

treatment_lum_summaries_part3 = plate_vals_part3 %>% 
	group_by(compound, concentration_M, cell_line) %>%
	summarise(sd = sd(lum),
						num_samples = n(),
						mean_lum = mean(lum),
						CV = sd/mean_lum)

ggplot(treatment_lum_summaries_part2, aes(x=CV)) + geom_histogram()
ggplot(treatment_lum_summaries_part3, aes(x=CV)) + geom_histogram()

```

```{r}
plate_DMSO_means_part1 = plate_vals_part1 %>%
	filter(compound == "DMSO") %>%
	group_by(row, plate_name) %>%
	summarise(DMSO_mean = mean(lum),
						num_DMSO = n())

plate_DMSO_means_part2 = plate_vals_part2 %>%
	filter(compound == "DMSO") %>%
	group_by(row, plate_name) %>%
	summarise(DMSO_mean = mean(lum),
						num_DMSO = n())

plate_DMSO_means_part3 = plate_vals_part3 %>%
	filter(compound == "DMSO") %>%
	group_by(row, plate_name) %>%
	summarise(DMSO_mean = mean(lum),
						num_DMSO = n())

if (! all(plate_DMSO_means_part2$num_DMSO == max(plate_DMSO_means_part2$num_DMSO))) {
	stop("Number of DMSO values in a plate isn't correct.")
} else {
	plate_DMSO_means_part2 = plate_DMSO_means_part2 %>% select(-num_DMSO)
}

if (! all(plate_DMSO_means_part3$num_DMSO == max(plate_DMSO_means_part3$num_DMSO))) {
	stop("Number of DMSO values in a plate isn't correct.")
} else {
	plate_DMSO_means_part3 = plate_DMSO_means_part3 %>% select(-num_DMSO)
}
```

```{r}
qc_filter <- function(data_set, viability_threshold = 120) {
	qc_fails = data_set %>% 
		filter(viability > viability_threshold) %>% 
		select(row,compound,cell_line,plate_name) %>% 
		unique() %>%
		mutate(qc_fail = T)
	
	data_set = data_set %>%
		left_join(qc_fails) %>%
		mutate(qc_fail = ifelse(is.na(qc_fail), F, T))
	
	return(data_set)
}

dir.create(here('results/validation_results/'), showWarnings = F)

plate_vals_part1 = plate_vals_part1 %>%
	left_join(plate_DMSO_means_part1) %>% 
	mutate(viability = lum/DMSO_mean*100) %>%
	write_rds(here('results/validation_results/validation_screen_part1.rds'))

plate_vals_part2 = plate_vals_part2 %>%
	left_join(plate_DMSO_means_part2) %>% 
	mutate(viability = lum/DMSO_mean*100) %>%
	qc_filter() %>%
	write_rds(here('results/validation_results/validation_screen_part2.rds'))

plate_vals_part3 = plate_vals_part3 %>%
	left_join(plate_DMSO_means_part3) %>% 
	mutate(viability = lum/DMSO_mean*100) %>%
	qc_filter() %>%
	write_rds(here('results/validation_results/validation_screen_part3.rds'))

treatment_via_summaries_part2 = plate_vals_part2 %>% 
	group_by(compound, concentration_M, cell_line) %>%
	summarise(sd = sd(viability),
						num_samples = n(),
						mean_via = mean(viability),
						CV = sd/mean_via)
treatment_via_summaries_part3 = plate_vals_part3 %>% 
	group_by(compound, concentration_M, cell_line) %>%
	summarise(sd = sd(viability),
						num_samples = n(),
						mean_via = mean(viability),
						CV = sd/mean_via)
```

```{r}
a = ggplot(plate_vals_part1 %>% filter(compound != "DMSO"), aes(x=concentration_M, y = viability)) +
	geom_hline(aes(yintercept = 90),alpha=0.5) +
	scale_x_log10() +
	geom_jitter() +
	geom_smooth() +
	BerginskiRMisc::theme_berginski() +
	facet_grid(rows = vars(cell_line), cols = vars(compound)) +
	ggtitle("Replicate 1")

b = ggplot(plate_vals_part2 %>% filter(compound != "DMSO"), aes(x=concentration_M, y = viability, color=qc_fail)) +
	geom_hline(aes(yintercept = 90),alpha=0.5) +
	geom_hline(aes(yintercept = 100),alpha=0.5, linetype = 2) +
	geom_hline(aes(yintercept = 130),alpha=0.5, linetype = 2) +
	scale_x_log10() +
	geom_jitter() +
	geom_smooth() +
	BerginskiRMisc::theme_berginski() +
	facet_grid(rows = vars(cell_line), cols = vars(compound)) +
	ggtitle("Replicate 2")

c = ggplot(plate_vals_part3 %>% filter(compound != "DMSO"), aes(x=concentration_M, y = viability, color=qc_fail)) +
	geom_hline(aes(yintercept = 90),alpha=0.5) +
	geom_hline(aes(yintercept = 100),alpha=0.5, linetype = 2) +
	geom_hline(aes(yintercept = 130),alpha=0.5, linetype = 2) +
	scale_x_log10() +
	geom_jitter() +
	geom_smooth() +
	BerginskiRMisc::theme_berginski() +
	facet_grid(rows = vars(cell_line), cols = vars(compound)) +
	ggtitle("Replicate 3")

d = b / c 

ggsave(here('figures/validation_testing/validation_viability_part2_part3.png'),width=20*0.75,height=15*0.75)
BerginskiRMisc::trimImage(here('figures/validation_testing/validation_viability_part2_part3.png'))
```